Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2022-12-21 manuscript.tex
Edits to abstract, introduction and discussion

\title{Different rules for binocular combination of luminance flicker in cortical and subcortical pathways}
\author{Federico G. Segala, Aurelio Bruno, Myat T. Aung, Alex R. Wade \& Daniel H. Baker}
\date{2022-10-16}

@@

@@

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

OK, some pointers for the introduction:

I think we need to draw a distinction between binocular summation at threshold (on which there is a big literature) and binocular facilitation above threshold. Much of the introduction is focussed on threshold measures, but actually I think our focus here is more about suprathreshold conditions where we usually see little facilitation. I think it will help the reader if we use the term summation only to refer to threshold phenomena, and facilitation in the more general case (I've tried to do this elsewhere in the paper).

The key points should be:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  For grating stimuli, there are summation effects at threshold, but these are lost at high contrasts.
\item
  Binocular combination happens in multiple pathways, including the pupil pathway. (This is a good place to mention the Quaia paper).
\item
  We don't know much about the pupil response, though it is clearly binocular because of the consensual response (convergence and divergence seem to be other ways of saying the same thing, so not sure they are worth mentioning).
\item
  Actually there's not much work on temporal binocular combination even in the primary visual pathway.
\item
  We have developed a novel paradigm that allows us to probe both cortical and subcortical pathways simultaneously for flickering light.
\end{enumerate}

I wouldn't say too much about models, or about the matching experiment, at this stage.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Binocular combination provides a higher visual sensitivity than monocular viewing. This superiority is known as binocular summation and is defined by the binocular summation ratio (BSR), which was originally believed to be around \(\sqrt{2}\) (\(\approx1.4\)) for grating stimuli at detection threshold (Campbell and Green, 1965). In other words, a monocular stimulus can elicit the same response as a binocular stimulus if it has a contrast that is 1.4 times higher. This led Legge to develop a widely accepted explanation that used quadratic summation to describe binocular combination (Legge, 1984): monocular signals from the right (R) and the left (L) eyes are squared before being summed together and the binocular response (B) is given by the square root of the output (B = \(\sqrt{R^2 + L^2}\), when R and L are equal to 1, the output is \(\sqrt{2}\)). However, subsequent research has shown that these explanations are not fully adequate to account for binocular summation. Both of these accounts constitute single channel models, and this type of model has been shown to not being able to account for contrast detection in the presence of noise (Anderson and Movshon, 1989). Moreover, more recent research has shown that the summation ratio can vary greatly between \(\sqrt{2}\) and 2 depending on factors such as the spatial and temporal frequency of a stimulus or the sensitivity difference between the eyes (Baker et al., 2018). These observations led to the development of multistage gain control models, which combine binocular summation and interocular suppression, and can account for contrast matching, detection and discrimination for spatial contrast (Ding and Sperling, 2006; Meese et al., 2006).

In general, it seems that the mechanisms behind binocular combination have been thoroughly studied, as have been the anatomical pathways behind it: light enters the eye though the pupil and signals are sent from the left and right retinae to the primary visual cortex, remaining anatomically isolated while passing through the lateral geniculate nucleus (LGN) until they reach V1, where they are binocularly combined (Purves et al., 2008). However, there is an eye component that is often underestimated in its role to determine the quality of visual information: the pupil. The pupils are openings found in the centre of the eyes that appear to be black and allow light to enter the eyes. Their size determines how much light will reach the retina and it is usually determined by the ambient levels of light: in brightness the pupils will constrict and in darkness they will dilate. This is known as the pupillary light response (PLR). The anatomical pathways that regulate this response are well understood and are very clearly and extensively described in the literature (Angée et al., 2021; Mathôt, 2018; McDougal and Gamlin, 2010; Wang and Munoz, 2015). However, they are anatomically distinct from the LGN-V1 pathway meaning that binocular combination occurs separately in anatomically distinct pathways. Given this, not much is known about the computational processes behind the PLR except for some evidence of binocular interaction. The presence of a consensual response in one eye when the other is being stimulated, and the presence of convergence (one pupil responds to illumination in either retina) and divergence (both pupils respond to illumination of one retina) (Wyatt and Musselman, 1981) are evidence of this binocular interaction.

With this in mind we designed an experiment that simultaneously recorded electrophysiological and pupillometric responses to investigate the combination of flickering light signals in both the visual cortex and the pupils. The results should offer new insight about basic neural circuits and information on how they might be affected in clinical disorders of vision (e.g.~amblyopia). Based on previous literature, we expected to find a non-linear combination of the responses in visual cortex, as described by the gain control mechanisms, and a more linear combination of the responses and a greater binocular response at the level of the pupils.

Several models have been developed and proposed to describe binocular combination and contrast matching results for spatial contrast. Among these, the models proposed by Ding and Sperling (2006) and Meese and colleagues (2006) are widely accepted. These models incorporate a dynamic contrast gain control, which allows to develop a model that combines binocular summation and interocular suppression The first model is a multistage model that consists of two pairs of gain control mechanisms that inhibit each other (Ding and Sperling, 2006). In the earlier stage, the right and left eye channels exert gain control on the other channel. Then, they exert gain control on the other channel's gain control. Lastly, the outputs from each channel are summed binocularly to determine the binocular signal's magnitude. The second model is similar as it also incorporates gain control mechanisms (Meese et al., 2006). In the first stage, it is applied to the left and right eye channels with suppression from the other eye. Then the signal from the two eyes is summed binocularly and gain control is applied a second time. This second model in particular is interesting and relevant as it was designed to also explain data from matching experiments.

To follow up on the results that we obtained, we decided to perform a contrast matching experiment to investigate whether perception of flickering light is consistent with the results observed in the cortical pathways (visual cortex) or the subcortical pathways (pupils). Matching is a paradigm in which the perceived brightness of a standard stimulus is matched to that of a target stimulus. In the latter, the interocular ratios of luminance are varied to obtain an equibrightness curve. Previous literature has used this paradigm to investigate the binocular fusion of static stimuli and the temporal combination of spatial flickering of spatial increments (a bright target on a dark background) and decrements (a dark target on a bright background) (Anstis and Ho, 1998; Levelt, 1965). For spatial increments, it was found that binocular fusion seems to follow approximately linear combination rules. This means that, for a monocular stimulus to elicit the same response as a binocular stimulus, the former needs to have twice the signal of the latter. On the other hand, spatial decrements follow a winner-takes-all pattern. This means that the observer is seeing what the eye that is receiving the strongest signal is seeing.

Our experiment used different stimuli than the ones used by Anstis and Ho (1998): in the binocular fusion experiment, their stimuli were not flickering and, in the flicker experiment, they were always shown in both eyes. In our experiment, in some conditions, the flicker was shown to only one eye. Moreover, they were looking at the temporal fusion of the flicker while we focussed on the binocular fusion of the flicker. Based on this and on the results from our first experiment, we expected to find a near linear summation of the responses.

@@

All stimuli were displayed at a mean luminance of 42 cd/m\(^2\) on an Iiyama Vision Master™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 (Minolta Camera Co.~Ltd., Japan). For experiments 1 and 2, the stimuli were presented using Psychopy (v3.0.7). For experiment 3, the stimuli were presented using Psychopy (v2022.1.1).

@@

The experiment was conducted in a windowless room, in which the only light source was the monitor. The participants sat at 99 cm from the monitor and the total optical viewing distance (through the stereoscope) was 107 cm. The experiment was carried out in one session lasting 45 minutes in total, divided in three blocks of 15 minutes each. In each block, there were 60 trials in total lasting 15 seconds each (12s of stimulus presentation, with an interstimulus interval of 3s). The participants were given no specific task instructions, but were asked to look at the fixation crosses in the middle of the two discs while trying to minimise their blinking during the presentation period.
We included six distinct ocular conditions, each at five temporal target contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed luminance contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e.~the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

The experiment was conducted in a darkened room with a blacked-out window. The display equipment (monitor and stereoscope) were the same as for the two previous experiments, but no EEG or pupillometry data were collected. A two-interval contrast matching procedure was used to collect data. In one interval, participants were presented with a standard fused disc that flickered at a set contrast level (either 24 or 48\%), which was selected by the experimenter at the beginning of each block. In the other interval, a target disc flickering at varying contrast levels was displayed. The contrast level of the target was controlled by a 1-up, 1-down staircase moving in logarithmic (dB) steps of contrast. The ratio of flicker amplitudes in the left and right eyes was varied across blocks and was set to be 0, 0.25, 0.5, 0.75 or 1 (9 distinct conditions). The standard and target discs were displayed for 1 second each, with an interstimulus interval of 0.5 seconds. After the discs appeared on screen, the participants had to indicate which interval they perceived as having the more intense flicker. The intervals were randomly ordered, and all discs flickered at a frequency of 2 Hz (two cycles in sine phase).

Due to its long duration (approximately 3 hours in total), the participants completed the experiment across multiple sessions initiated at their own convenience The experiment was divided into 54 blocks (3 repetitions \(\times\) 2 standard contrasts \(\times\) 9 target ratios), which lasted on average 3 minutes each, depending on the response speed of the participant. In each block, there were a total of 50 trials. No auditory feedback was given for this subjective task.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox (Delorme and Makeig, 2004). The data for each participant were then loaded into R for analysis where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The fast Fourier transform was calculated for each waveform, and the spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging retained the phase information (i.e.~coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding \(D\) = 3 from the complex-valued mean (see Baker, 2021). For statistical comparisons of complex-valued data, we use the \(ANOVA^2_{circ}\) statistic described by Baker (2021). This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the \(T^2_{circ}\) statistic of Victor and Mast (1991) that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software (Kassner et al., 2014), which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The fast Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit sine-waves to the data, rather than using Fourier analysis (e.g. Spitschan et al., 2014). Our novel approach is more robust to noise at other frequencies, and makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis (e.g. Figueira et al., 2022)

To analyse the matching data, we pooled the trial responses across all repetitions of a given condition for each participant. We then fitted a cumulative normal psychometric function to estimate the point of subjective equality at the 50\% level. Thresholds were averaged across participants in logarithmic (dB) units. For all experiments, we used a bootstrapping procedure with 1000 iterations to estimate standard errors across participants. All analysis and figure construction was conducted using a single R-script, available online, making this study fully computationally reproducible.

@@

where \emph{n} is a noise parameter, and \(R_{max}\) scales the overall response amplitude. The \(R_{max}\) parameter was omitted when modelling the contrast matching data, as it had no effect in this paradigm.

Despite being derived from the model proposed by Meese et al. (2006), the simplifications applied to this architecture make it very similar to other models (e.g. Ding and Sperling, 2006; Legge, 1984; Schrödinger, 1926). In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression (see Baker et al., 2012; Kingdom and Libenson, 2015). Our key parameter of interest is the weight of interocular suppression. Large values around \(w\) = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli (Baker and Wade, 2017). Low values around \(w\) = 0 produce substantial, near-linear binocular facilitation (Baker et al., 2020).

We implemented the model within a Bayesian framework using the Stan software (Carpenter et al., 2017). This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, \(w\), and the other free model parameters \(R_{max}\), \(Z\) and \(n\). The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for \(w\) and \(R_{max}\), and 5 and 2 for \(Z\) and \(n\). We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data. We sampled over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

The pupillometry results are summarised in Figure \ref{fig:pupildata}. The group average waveform for binocular presentation is shown in Figure \ref{fig:pupildata}a. There is a substantial pupil constriction at stimulus onset, followed by visible oscillations at the flicker frequency (2Hz, see waveform at foot). The average Fourier spectrum is displayed in Figure \ref{fig:pupildata}b, and shows a substantial spike at 2 Hz, but no evidence of a second harmonic response at 4Hz. These results demonstrate that our paradigm can evoke measurable steady-state pupil responses at 2Hz.

@@

Figure \ref{fig:pupildata}e shows responses at 1.6Hz, for the same conditions, as well as for a condition in which a monocular stimulus flickered at 1.6Hz (grey circles). Surprisingly there appears to be a slight facilitation effect in the binocular cross condition, particularly at lower contrasts. The dichoptic cross condition does not show clear modulation with target contrast.

@@

The strong binocular facilitation and weak interocular suppression in the EEG data from Experiment 1 was very different from previous findings on binocular combination using steady-state EEG with grating stimuli (Baker and Wade, 2017). One possible explanation is that the lower temporal frequency used here (2Hz, vs 5 or 7Hz in previous work) might be responsible for this difference. We therefore ran a second experiment to compare monocular and binocular responses at a range of temporal frequencies. Only EEG data were collected for this experiment, as the pupil response is negligible above around 2Hz (Spitschan et al., 2014).

Results from the temporal frequency experiment are shown in \ref{fig:TFdata}. Figure \ref{fig:TFdata}a shows the Fourier spectra for responses to binocular flicker at 5 different frequencies (2, 4, 8, 16, and 30 Hz). From 2 to 16 Hz, clear signals are observed at each fundamental frequency, and typically also their higher harmonics (integer multiples of the fundamental). However, at 30 Hz (upper row), the responses recorded were not demonstrably above the noise baseline. Figure \ref{fig:TFdata}b compares the monocular and binocular responses at each stimulation frequency. Here we replicate the substantial summation effect across frequencies up to and including 16Hz (Fig. \ref{fig:TFdata}c), demonstrating that strong binocular facilitation in the EEG data of Experiment 1 cannot be attributed to our use of 2Hz flicker.

@@

In Experiment 1 we found evidence of much stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for percieved contrast judgements indexed by psychophysical contrast matching paradigms (e.g. Anstis and Ho, 1998; Levelt, 1965; Quaia et al., 2018). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

>>>>

\title{Different rules for binocular combination of luminance flicker in cortical and subcortical pathways}
\author{Federico G. Segala, Aurelio Bruno, Myat T. Aung, Alex R. Wade \& Daniel H. Baker}
\date{2022-12-21}

@@

How does the brain combine information across the eyes? In primary visual cortex, spatial patterns are subject to `ocularity invariance', where monocular and binocular responses are rendered equal by mutual suppression between the eyes. Here we find a very different pattern of activity in response to luminance flicker, both in the cortex and also in the subcortical pathway that determines pupil diameter, by simultaneously collecting steady-state EEG and pupillometry data. We find strong binocular facilitation, and negligible interocular suppression from cortical sources, whereas pupil diameters showed weaker facilitation and stronger suppression. We confirm that this is not a consequence of the temporal parameters chosen, as pronounced binocular facilitation was also observed at faster flicker rates. Near-linear binocular combination was also found for the same stimuli using a matching task. A hierarchical Bayesian implementation of a standard binocular combination model confirms that interocular suppression was substantially weaker for our EEG and matching data, compared with the pupillometry results. These findings illustrate that the brain can repurpose a generic algorithm for different functions by adjusting parameters such as the weight of suppression between pathways.

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within (Baker and Wade, 2017) and between (Ernst and Banks, 2002) the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established (Hubel and Wiesel, 1962). Yet much of our knowledge about binocular combination focuses on the `canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) (Purves et al., 2008). Signals are also combined across the eyes in a network of subcortical nuclei that govern pupil diameter in response to incoming light levels (McDougal and Gamlin, 2008). Our primary purpose here is to investigate whether signal combination involves the same or different operations across these two anatomically distinct pathways.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation, with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between \(\sqrt{2}\) and 2 (Baker et al., 2018; Campbell and Green, 1965). But this advantage is lost at high stimulus intensities, where both psychophysical performance (contrast discrimination thresholds) (Legge, 1984; Meese et al., 2006) and neural activity (Baker and Wade, 2017; Moradi and Heeger, 2009) are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision (Ding and Sperling, 2006; Meese et al., 2006) advocate a process of interocular suppression that, at high contrasts, normalizes the two eyes inputs and negates the binocular advantage. This is consistent with our everyday experience of `ocularity invariance' (Baker et al., 2007): perceived contrast does not change when one eye is opened and closed.

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (McDougal and Gamlin, 2008). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (1981)). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (Angée et al., 2021; Mathôt, 2018; McDougal and Gamlin, 2008; Wang and Munoz, 2015). To account for the consensual response, these brain regions must necessarily combine information from the left and right eyes (Doesschate and Alpern, 1967), yet the precise algorithm by which this is achieved is currently unclear.

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. This novel paradigm allows us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Using periodic flicker entrains both neural (Norcia et al., 2015) and pupil (Spitschan et al., 2014) responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. We follow up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e.~temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways involve stronger interocular suppression than the canonical cortical pathway.

@@

All stimuli had a mean luminance of 42 cd/m\(^2\) on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co.~Ltd., Japan). For experiments 1 and 2, the stimuli were presented using Psychopy (v3.0.7). For experiment 3, the stimuli were presented using Psychopy (v2022.1.1).

@@

The experiment was conducted in a windowless room, in which the only light source was the monitor. The participants sat at 99 cm from the monitor and the total optical viewing distance (through the stereoscope) was 107 cm. The experiment was carried out in one session lasting 45 minutes in total, divided in three blocks of 15 minutes each. In each block, there were 60 trials in total lasting 15 seconds each (12s of stimulus presentation, with an interstimulus interval of 3s). The participants were given no specific task instructions, but were asked to look at the fixation cross in the middle of the disc while trying to minimise their blinking during the presentation period.

We included six distinct ocular conditions, each at five temporal target contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast is defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e.~the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

The experiment was conducted in a darkened room with a blacked-out window. The display equipment (monitor and stereoscope) were the same as for the two previous experiments, but no EEG or pupillometry data were collected. A two-interval contrast matching procedure was used to collect data. In one interval, participants were presented with a standard fused disc that flickered at a set contrast level (either 24 or 48\%), which was selected by the experimenter at the beginning of each block. In the other interval, a target disc flickering at varying contrast levels was displayed. The contrast level of the target was controlled by a 1-up, 1-down staircase moving in logarithmic (dB) steps of contrast. The ratio of flicker amplitudes in the left and right eyes was varied across blocks and was set to be 0, 0.25, 0.5, 0.75 or 1 (9 distinct conditions). The standard and target discs were displayed for 1 second each, with an interstimulus interval of 0.5 seconds. After both discs had appeared on screen, the participants had to indicate which interval they perceived as having the more intense flicker. The intervals were randomly ordered, and all discs flickered at a frequency of 2 Hz (two cycles in sine phase).

Due to its long duration (approximately 3 hours in total), the participants completed the experiment across multiple sessions initiated at their own convenience. The experiment was divided into 54 blocks (3 repetitions \(\times\) 2 standard contrasts \(\times\) 9 target ratios), which lasted on average 3 minutes each, depending on the response speed of the participant. In each block, there were a total of 50 trials. No auditory feedback was given for this subjective task.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox (Delorme and Makeig, 2004). The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The fast Fourier transform was calculated for each waveform, and the spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging retained the phase information (i.e.~coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding \(D\) = 3 from the complex-valued mean (see Baker, 2021). For statistical comparisons of complex-valued data, we use the \(ANOVA^2_{circ}\) statistic described by Baker (2021). This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the \(T^2_{circ}\) statistic of Victor and Mast (1991) that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software (Kassner et al., 2014), which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The fast Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit sine-waves to the data, rather than using Fourier analysis (e.g. Spitschan et al., 2014). The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies (see Barrionuevo et al., 2014; Barrionuevo and Cao, 2016). Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis (e.g. Figueira et al., 2022).

To analyse the matching data, we pooled the trial responses across all repetitions of a given condition for each participant. We then fitted a cumulative normal psychometric function to estimate the point of subjective equality at the 50\% level. Thresholds were averaged across participants in logarithmic (dB) units.

For all experiments, we used a bootstrapping procedure with 1000 iterations to estimate standard errors across participants. All analysis and figure construction was conducted using a single R-script, available online, making this study fully computationally reproducible.

@@

where \emph{n} is a noise parameter, and \(R_{max}\) scales the overall response amplitude. The \(R_{max}\) parameter was omitted when modelling the contrast matching data, as it has no effect in this paradigm.

Despite being derived from the model proposed by Meese et al. (2006), the simplifications applied to this architecture make it very similar to other models (e.g. Ding and Sperling, 2006; Doesschate and Alpern, 1967; Legge, 1984; Schrödinger, 1926). In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression (see Baker et al., 2012; Kingdom and Libenson, 2015). Our key parameter of interest is the weight of interocular suppression. Large values around \(w\) = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli (Baker and Wade, 2017). Low values around \(w\) = 0 produce substantial, near-linear binocular facilitation (Baker et al., 2020).

We implemented the model within a Bayesian framework using the Stan software (Carpenter et al., 2017). This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, \(w\), and the other free model parameters \(R_{max}\), \(Z\) and \(n\). The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for \(w\) and \(R_{max}\), and 5 and 2 for \(Z\) and \(n\). We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data. We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

The pupillometry results are summarised in Figure \ref{fig:pupildata}. The group average waveform for binocular presentation is shown in Figure \ref{fig:pupildata}a. There is a substantial pupil constriction at stimulus onset, followed by visible oscillations at the flicker frequency (2Hz, see waveform at foot). The average Fourier spectrum is displayed in Figure \ref{fig:pupildata}b, and shows a clear spike at 2 Hz, but no evidence of a second harmonic response at 4Hz. These results demonstrate that our paradigm can evoke measurable steady-state pupil responses at 2Hz.

@@

Figure \ref{fig:pupildata}e shows responses at 1.6Hz, for the same conditions, as well as for a condition in which a monocular stimulus flickered at 1.6Hz (grey circles). Surprisingly there again appears to be a slight facilitation effect in the binocular cross condition, particularly at lower contrasts. The dichoptic cross condition does not show clear modulation with target contrast.

@@

The strong binocular facilitation and weak interocular suppression in the EEG data from Experiment 1 was very different from previous findings on binocular combination using steady-state EEG with grating stimuli (Baker and Wade, 2017). One possible explanation is that the lower temporal frequency used here (2Hz, vs 5 or 7Hz in previous work) might be responsible for this difference. We therefore ran a second experiment to compare monocular and binocular responses at a range of temporal frequencies. Only EEG data were collected for this experiment, as the pupil response is negligible above around 2Hz (Spitschan et al., 2014); note that we originally chose 2Hz because it produces measurable signals for both EEG and pupillometry, yet is unfortunately optimal for neither.

Results from the temporal frequency experiment are shown in Figure \ref{fig:TFdata}. Figure \ref{fig:TFdata}a shows the Fourier spectra for responses to binocular flicker at 5 different frequencies (2, 4, 8, 16, and 30 Hz). From 2 to 16 Hz, clear signals are observed at each fundamental frequency, and typically also their higher harmonics (integer multiples of the fundamental). However, at 30 Hz (upper row), the responses recorded were not demonstrably above the noise baseline. Figure \ref{fig:TFdata}b compares the monocular and binocular responses at each stimulation frequency. Here we replicate the substantial summation effect across frequencies up to and including 16Hz (Fig. \ref{fig:TFdata}c), demonstrating that strong binocular facilitation in the EEG data of Experiment 1 cannot be attributed to our use of 2Hz flicker.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. Anstis and Ho, 1998; Levelt, 1965; Quaia et al., 2018). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

Using a novel paradigm that combines EEG and pupillometry, we found surprising results for the binocular combination of flickering light. In the visual cortex, the processing of temporal luminance modulations seems to happen approximately linearly, with no evidence of interocular suppression. In the subcortical pathway, signal combination is more non-linear, with evidence of interocular suppression. This pattern of results was also supported by computational modelling, which showed a much greater suppressive weight for the pupillometry data compared to the EEG data. Additionally, we found that perception of flickering light is consistent with a near-linear summation process, consistent with the cortical (EEG) responses.

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e.~sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation (Baker and Wade, 2017; Meese et al., 2006; Moradi and Heeger, 2009). Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al. (2018) observed a strong binocular facilitation (or `supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and Cajochen (2019) report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm (Baker et al., 2020). Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments (Anstis and Ho, 1998; Baker et al., 2012; Levelt, 1965), though not for luminance decrements (Anstis and Ho, 1998). Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? Baker and Wade (2017) point out that strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs (see also Ernst and Banks, 2002). This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction (Li and Atick, 1994; May and Zhaoping, 2022). One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or otherwise be phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.
